---
title: "The Poorest School Districts in America"
author: "Richard Paquin Morel"
date: "3/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(educationdata)
library(tigris)
library(sf)
options(tigris_use_cache = TRUE)

theme_set(
  theme_minimal() +
    theme(
      text = element_text(family = "Palatino"),
      axis.title.x = element_text(size = 10, hjust = 1),
      axis.title.y = element_text(size = 10),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(size = 10, colour = "grey25", face = "bold"),
      plot.subtitle = element_text(size = 9, colour = "grey45"),
      legend.position = "bottom", 
      legend.box="vertical",
      legend.margin=margin(),
      legend.text = element_text(size = 6), 
      legend.title = element_text(size = 8, vjust = 0),
      legend.key = unit(0.5, "cm")
      )
  )
```

## School Funding in America, a Preamble

It's no secret that there are wide disparities in the funding of schools in the United States. Constitutionally, education is a reserved power of the states, so the federal government is limited, policy-wise, in how it can intervene in educational matters. Since the Great Society legislation of the 1960s, the federal role in education has been supplemental funding for schools serving low income students. Culturally, there is a strong tradition of "local control" over schools--schools are administered and funded at the district level. Most school funding comes from state and local sources. How involved state governments are vary greatly from state to state.

```{r get data}
district_funding <-
  get_education_data(level = "school-districts", 
                     source = "ccd",
                     topic = "finance",
                     filters = list(year = 2015)) %>% 
  na_if(-1) %>% 
  na_if(-2) %>% 
  na_if(-3) %>% 
  mutate(leaid = as.character(as.numeric(leaid)))

d_dir <-
  get_education_data(level = "school-districts", 
                     source = "ccd",
                     topic = "directory",
                     filters = list(year = 2015)) %>% 
  na_if(-1) %>% 
  na_if(-2) %>% 
  na_if(-3) %>% 
  mutate(leaid = as.character(as.numeric(leaid)))

d_enrollment <-
  get_education_data(level = "school-districts", 
                     source = "ccd", 
                     topic = "enrollment", 
                     by = "race", 
                     filters = list(year = 2015, grade = 99)) %>% 
  mutate(
    race = case_when(
        race == 1 ~ "white",
        race == 2 ~ "black",
        race == 3 ~ "hispanic",
        race == 4 ~ "asian",
        race == 5 ~ "other",
        race == 6 ~ "other",
        race == 7 ~ "other",
        race == 8 ~ "other", 
        race == 9 ~ "other",
        race == 20 ~ "other",
        race == 99 ~ "total"
    )
) %>% 
    pivot_wider(id_cols = c(leaid, year), 
                names_from = race, 
                values_from = enrollment,
                values_fn = list(enrollment = ~(sum(., na.rm = TRUE)))) %>% 
  mutate(leaid = as.character(as.numeric(leaid)),
         nonwhite = total - white) %>% 
  mutate_at(vars(white:other, nonwhite), list(prop = ~ ./total))

fips <- fips_codes %>% select(state, state_code) %>% distinct() %>% filter(state_code <= 56) %>% pull(state_code)

district_geo <-
  map(fips, ~ school_districts(.x, year = 2015, class = "sf"))

district_geo <- reduce(district_geo, rbind)
#B19013_001
median_income <- map(fips, ~ 
                       get_acs(geography = "school district (unified)", 
                               variables = "B19013_001", 
                               year = 2015,
                               state = .x)
                     ) %>% 
  bind_rows() %>% 
  select(-NAME)

district_geo <- left_join(district_geo, median_income, by = "GEOID")

district_funding <- 
  district_funding %>% 
  left_join(d_dir %>% select(leaid, lea_name, city_location, state_location)) %>% 
  left_join()
  inner_join(district_geo %>%
               select(GEOID, estimate, geometry) %>% 
               mutate(GEOID = as.character(as.numeric(GEOID))), 
             by = c("leaid" = "GEOID"))
```

## The Landscape of school funding 

Where does school funding come from?

```{r}
district_funding %>% 
  drop_na(state_location) %>% 
  select(fips, leaid, state_location,
         rev_fed_total, 
         rev_state_total, rev_local_total,
         enrollment_fall_responsible,
         estimate) %>% 
  group_by(state_location) %>% 
  summarize_at(vars(contains("rev")), sum, na.rm = TRUE) %>% 
  pivot_longer(cols = c(-state_location)) %>% 
  group_by(state_location) %>% 
  mutate(prop_value = value/sum(value, na.rm = TRUE)) %>% 
  ggplot(aes(x = state_location, y = prop_value, group = name)) +
  geom_col(aes(fill = name), alpha = 0.7) +
  scale_fill_viridis_d(name = "Revenue Source", labels = c("Federal", "Local", "State")) +
  labs(x = "", y = "Share", title = "Revenue Source by State",
       caption = "Source: My calculations using data from the Common Core of Data, accessed through the Urban Institute's API.") +
  coord_flip()
```


```{r revenue}
district_funding %>% 
  select(fips, leaid, city_location,
         rev_total, rev_fed_total, 
         rev_state_total, rev_local_total,
         enrollment_fall_responsible,
         estimate, nonwhite_prop) %>% 
  filter(rev_total != 0,
         rev_fed_total != 0,
         rev_state_total != 0,
         rev_local_total != 0,
         enrollment_fall_responsible != 0) %>% 
  drop_na(estimate) %>% 
  mutate_at(vars(enrollment_fall_responsible, contains("rev")), log) %>%
  mutate(income_ntile = ntile(estimate, 5)) %>% 
  pivot_longer(cols = c(-fips, -leaid, -city_location, -enrollment_fall_responsible, -income_ntile, -estimate, -nonwhite_prop)) %>% 
  mutate(name = case_when(
    name == "rev_fed_total" ~ "Total federal revenue",
    name == "rev_state_total" ~ "Total state revenue",
    name == "rev_local_total" ~ "Total local revenue",
    name == "rev_total" ~ "Total revenue",
  )) %>% 
  mutate(name = factor(name, levels = c("Total federal revenue",
                                        "Total state revenue",
                                        "Total local revenue",
                                        "Total revenue"))) %>% 
   ggplot(aes(x = enrollment_fall_responsible, y = value)) +
  geom_point(aes(color = factor(income_ntile)), alpha = 0.5) +
  geom_smooth(method = "lm", color = "plum4") +
  scale_color_viridis_d(name = "Income quantile") +
  labs(x = "Enrollment", y = "") +
  facet_wrap(~ name)
```

```{r}
district_funding %>% 
  select(fips, leaid, city_location,
         rev_total, rev_fed_total, 
         rev_state_total, rev_local_total,
         enrollment_fall_responsible,
         estimate, nonwhite_prop) %>% 
  filter(rev_total != 0,
         rev_fed_total != 0,
         rev_state_total != 0,
         rev_local_total != 0,
         enrollment_fall_responsible != 0) %>% 
  drop_na(estimate) %>% 
  mutate_at(vars(enrollment_fall_responsible, contains("rev")), log) %>%
  mutate(income_ntile = ntile(estimate, 5)) %>% 
  pivot_longer(cols = c(-fips, -leaid, -city_location, -enrollment_fall_responsible, -income_ntile, -estimate, -nonwhite_prop)) %>% 
  mutate(name = case_when(
    name == "rev_fed_total" ~ "Total federal revenue",
    name == "rev_state_total" ~ "Total state revenue",
    name == "rev_local_total" ~ "Total local revenue",
    name == "rev_total" ~ "Total revenue",
  )) %>% 
  mutate(name = factor(name, levels = c("Total federal revenue",
                                        "Total state revenue",
                                        "Total local revenue",
                                        "Total revenue"))) %>% 
   ggplot(aes(x = enrollment_fall_responsible, y = value)) +
  geom_point(aes(color = factor(income_ntile), size = nonwhite_prop), alpha = 0.5) +
  geom_smooth(method = "lm", color = "plum4") +
  scale_color_viridis_d(name = "Income quantile") +
  scale_size(name = "Share of students of color",
             range = c(0.1, 5)) +
  labs(x = "Enrollment", y = "",
       title = "Revenue source and enrollement",
       caption = "Source: My calculations using data from the Common Core of Data, accessed through the Urban Institute's API.") +
  facet_wrap(~ name)
```

```{r}
district_funding %>% 
  select(fips, leaid, city_location,
         rev_total, rev_fed_total, 
         rev_state_total, rev_local_total,
         enrollment_fall_responsible,
         estimate, nonwhite_prop, nonwhite) %>% 
  filter(rev_total != 0,
         rev_fed_total != 0,
         rev_state_total != 0,
         rev_local_total != 0,
         enrollment_fall_responsible != 0) %>% 
  drop_na(estimate) %>% 
  mutate_at(vars(enrollment_fall_responsible, contains("rev")), log) %>%
  mutate(income_ntile = ntile(estimate, 5)) %>% 
  pivot_longer(cols = c(-fips, -leaid, -city_location, 
                        -enrollment_fall_responsible, -income_ntile, 
                        -estimate, -nonwhite_prop, -nonwhite)) %>% 
  mutate(name = case_when(
    name == "rev_fed_total" ~ "Total federal revenue",
    name == "rev_state_total" ~ "Total state revenue",
    name == "rev_local_total" ~ "Total local revenue",
    name == "rev_total" ~ "Total revenue",
  )) %>% 
  mutate(name = factor(name, levels = c("Total federal revenue",
                                        "Total state revenue",
                                        "Total local revenue",
                                        "Total revenue"))) %>% 
   ggplot(aes(x = nonwhite, y = value)) +
  geom_point(aes(color = factor(income_ntile), size = enrollment_fall_responsible), alpha = 0.5) +
  geom_smooth(method = "lm", color = "plum4") +
  scale_color_viridis_d(name = "Income quantile") +
  scale_size(name = "Share of students of color",
             range = c(0.1, 5)) +
  labs(x = "Enrollment", y = "",
       title = "Revenue source and enrollement",
       caption = "Source: My calculations using data from the Common Core of Data, accessed through the Urban Institute's API.") +
  facet_wrap(~ name)
```

```{r}
district_funding %>% 
    select(fips, leaid, city_location,
           rev_total, rev_fed_total, 
           rev_state_total, rev_local_total,
           enrollment_fall_responsible,
           estimate, nonwhite_prop, nonwhite) %>% 
    filter(rev_total != 0,
           rev_fed_total != 0,
           rev_state_total != 0,
           rev_local_total != 0,
           enrollment_fall_responsible != 0) %>% 
    drop_na(estimate) %>% 
    mutate_at(vars(enrollment_fall_responsible, estimate, contains("rev")), log) %>%
    pivot_longer(cols = c(-fips, -leaid, -city_location, 
                          -enrollment_fall_responsible,
                          -estimate, -nonwhite_prop, -nonwhite)) %>% 
    mutate(name = case_when(
        name == "rev_fed_total" ~ "Total federal revenue",
        name == "rev_state_total" ~ "Total state revenue",
        name == "rev_local_total" ~ "Total local revenue",
        name == "rev_total" ~ "Total revenue",
    )) %>% 
    mutate(name = factor(name, levels = c("Total federal revenue",
                                          "Total state revenue",
                                          "Total local revenue",
                                          "Total revenue"))) %>% 
    ggplot(aes(x = estimate, y = value)) +
    geom_point(aes(color = nonwhite_prop), alpha = 0.5) + geom_smooth(method = "lm", color = "plum4") +
    scale_color_viridis_c(name = "Share of students of color") +
    labs(x = "Log of Median Income", y = "",
         title = "Revenue source and district median income",
         caption = "Source: My calculations using data from the Common Core of Data, accessed through the Urban Institute's API.") +
    facet_wrap(~ name)
```

## Per student funding from various sources

```{r}
district_funding %>% 
    select(fips, leaid, city_location,
           rev_total, rev_fed_total, 
           rev_state_total, rev_local_total,
           enrollment_fall_responsible,
           estimate, nonwhite_prop, nonwhite) %>% 
    filter(rev_total != 0,
           rev_fed_total != 0,
           rev_state_total != 0,
           rev_local_total != 0,
           enrollment_fall_responsible != 0) %>% 
    drop_na(estimate) %>% 
    mutate_at(vars(contains("rev")), 
              list(per_student = ~ ./enrollment_fall_responsible)) %>% 
    select(-ends_with("total")) %>% 
    mutate_at(vars(enrollment_fall_responsible, estimate, contains("per_student")), log) %>%
    pivot_longer(cols = c(-fips, -leaid, -city_location, 
                          -enrollment_fall_responsible,
                          -estimate, -nonwhite_prop, -nonwhite)) %>% 
    mutate(name = case_when(
        name == "rev_fed_total_per_student" ~ "Total per student federal revenue",
        name == "rev_state_total_per_student" ~ "Total per student state revenue",
        name == "rev_local_total_per_student" ~ "Total per student local revenue",
        name == "rev_total_per_student" ~ "Total per student revenue",
    )) %>% 
    mutate(name = factor(name, levels = c("Total per student federal revenue",
                                          "Total per student state revenue",
                                          "Total per student local revenue",
                                          "Total per student revenue"))) %>% 
    ggplot(aes(x = nonwhite_prop, y = value)) +
    geom_point(aes(color = estimate), alpha = 0.5) +
    geom_smooth(method = "lm", color = "plum4") +
    scale_color_viridis_c(name = "Logged Median Income") +
    labs(x = "Share of students of color", y = "",
         title = "Per student revenue by share of students of color",
         caption = "Source: My calculations using data from the Common Core of Data, accessed through the Urban Institute's API.") +
    facet_wrap(~ name, scales = "free_y")
```

